---
title: "Measuring AI involvement in a repository: the epiworld case"
format:
  html:
    toc: true
    fig-width: 9
    fig-height: 4.5
vignette: >
  %\VignetteIndexEntry{Measuring AI involvement in a repository: the epiworld case}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

This vignette walks through a complete analysis of AI involvement in a real
project: [epiworld](https://github.com/UofUEpiBio/epiworld), a C++
header-only library for agent-based epidemiological simulations developed at
the University of Utah. We will answer three questions:

1. **When** did AI coding agents become involved in the project?
2. **How much** of the commit and coding activity involves AI?
3. **How** did activity change after AI became involved?

```{r}
#| label: setup
library(aitracking)
library(data.table)
```

## Getting the data

Two calls retrieve everything we need. `gh_commits()` downloads the full
commit history, `gh_commit_lines()` adds per-commit line counts (one API call
per commit), and `gh_interactions()` downloads the issue/PR conversation
layer--which survives even when PRs are squash-merged:

```{r}
#| label: fetch
#| eval: false
commits <- gh_commits("UofUEpiBio/epiworld") |>
  gh_commit_lines()

interactions <- gh_interactions("UofUEpiBio/epiworld")
```

Both functions authenticate via `gh_token()` (an explicit token, the
`GITHUB_PAT`/`GITHUB_TOKEN` environment variables, or the git credential
store). So that this vignette builds offline, the package ships a snapshot of
both datasets taken on 2026-07-31, and we use it from here on:

```{r}
#| label: load-data
data(epiworld_commits)
data(epiworld_interactions)

commits      <- epiworld_commits
interactions <- epiworld_interactions

commits[, .(sha = substr(sha, 1, 7), author, date, additions, deletions)] |>
  head()
```

## When did AI become involved?

`ai_classify()` tags each commit using a small set of transparent rules: AI
identities among authors/committers (e.g., `copilot[bot]`), and attribution
trailers in the commit message (e.g., `Co-authored-by: Claude`). The default
patterns are in `ai_patterns()`; the generic `bot` pattern also catches
non-AI automation such as Dependabot, so we drop it here to focus on actual
AI coding agents:

```{r}
#| label: classify
ai_only <- ai_patterns()[names(ai_patterns()) != "bot"]

commits <- ai_classify(commits, patterns = ai_only)

# How many commits involved AI, and which agents?
commits[, .N, by = .(ai, ai_agent)]
```

```{r}
#| label: first-ai
first_ai <- commits[ai == TRUE, min(date)]
first_ai
```

So the first AI-involved commit happened on
`r format(first_ai, "%B %d, %Y")`:

```{r}
#| label: first-ai-commit
commits[ai == TRUE][which.min(date), .(
  sha = substr(sha, 1, 7), author, date,
  message = substr(message, 1, 60), ai_agent
)]
```

Note that this commit's *author* is a human. It was flagged because the
message carries an attribution trailer left by a squash-merged pull request:

```{r}
#| label: first-ai-trailer
commits[ai == TRUE][which.min(date), message] |>
  strsplit("\n") |>
  unlist() |>
  grep(pattern = "Co-authored-by", value = TRUE) |>
  unique()
```

This is the main reason `ai_classify()` looks at commit messages and not just
at author identities: when a PR written with an agent is squash-merged, the
resulting commit is attributed to whoever merged it, and the trailer is the
only trace of AI involvement left in the commit history.

## Visualizing the timeline

`plot_timeline()` shows monthly activity. Since the data went through
`ai_classify()`, commit bars are split into human and AI-involved:

```{r}
#| label: timeline-commits
plot_timeline(commits, by = "commits")
```

The same timeline in terms of lines added and deleted:

```{r}
#| label: timeline-lines
plot_timeline(commits, by = "lines")
```

## How did activity change since AI got involved?

A simple before/after comparison of monthly activity:

```{r}
#| label: before-after
commits[, period := fifelse(date < first_ai, "1. Before AI", "2. Since AI")]

stats <- commits[, .(
  months        = uniqueN(format(date, "%Y-%m")),
  commits       = .N,
  authors       = uniqueN(author_name),
  lines_added   = sum(additions, na.rm = TRUE),
  lines_deleted = sum(deletions, na.rm = TRUE)
), by = period]

stats[, `:=`(
  commits_month = round(commits / months, 1),
  added_month   = round(lines_added / months, 1),
  deleted_month = round(lines_deleted / months, 1)
)]

stats
```

And within the AI period, how much of the work is AI-involved?

```{r}
#| label: ai-share
share <- commits[period == "2. Since AI", .(
  commits     = .N,
  lines_added = sum(additions, na.rm = TRUE)
), by = ai]

share
```

```{r}
#| label: share-values
#| include: false
pc_commits <- round(100 * share[ai == TRUE, commits] / sum(share$commits))
pc_lines   <- round(100 * share[ai == TRUE, lines_added] / sum(share$lines_added))
rate_before <- stats[period == "1. Before AI", commits_month]
rate_since  <- stats[period == "2. Since AI", commits_month]
```

Two things stand out. First, AI involvement is substantial once it starts:
`r pc_commits`% of commits and `r pc_lines`% of added lines in the AI period
are AI-involved. Second--and this is why the raw counts deserve care--overall
activity did **not** accelerate: the project went from `r rate_before` to
`r rate_since` commits per month.

That drop is a good reminder that this is observational data, not an
experiment. epiworld is a mature library, and the AI period coincides with a
phase of consolidation rather than initial development; the same comparison
on a young project would likely look very different. What the data supports
is a statement about *composition* (how much of the work AI touches), not
about causation (whether AI made the project faster).

## Human requests to AI: the conversation layer

Commit histories under-count AI involvement: when a PR authored by an agent
is squash-merged, the merge commit may carry a human identity. The
issue/PR conversation retrieved by `gh_interactions()` preserves that
history. `ai_classify()` adds two flags here: `ai` (the comment/issue was
authored by an AI agent) and `ai_mention` (the text addresses or mentions an
agent--a proxy for humans prompting AI):

```{r}
#| label: interactions
interactions <- ai_classify(interactions, patterns = ai_only)

# Overall activity by type
interactions[, .N, by = type]

# Comments/issues authored by AI agents
interactions[ai == TRUE, .N, by = ai_agent]

# Humans prompting/addressing AI agents
interactions[ai == FALSE & !is.na(ai_mention), .N, by = ai_mention]
```

Activity by user (top 10), distinguishing AI agents from humans:

```{r}
#| label: by-user
interactions[, .N, by = .(user, ai)][order(-N)] |> head(10)
```

## Evolution of the project's size

`loc_evolution()` accumulates net lines changed (additions minus deletions)
over time--an approximation of the project's size in lines:

```{r}
#| label: loc
lo <- loc_evolution(commits)

plot(
  lo$date, lo$loc, type = "l", lwd = 2, col = "steelblue",
  xlab = "Date", ylab = "Cumulative net lines",
  main = "epiworld: project size over time"
)
abline(v = as.numeric(first_ai), lty = 2, col = "tomato")
text(
  as.numeric(first_ai), max(lo$loc) * 0.1, "first AI commit",
  pos = 4, col = "tomato"
)
```

For a by-language breakdown, retrieve file-level detail with
`gh_commit_lines(commits, files = TRUE)` and pass that to `loc_evolution()`,
or get the current snapshot with:

```{r}
#| label: languages
#| eval: false
gh_languages("UofUEpiBio/epiworld")
```

## Caveats

- `ai_classify()` is deliberately parsimonious: it only catches AI
  involvement that is *visible* (bot identities, attribution trailers,
  mentions). Undisclosed AI assistance--e.g., code written with editor
  completions and committed under a human identity--is not detectable from
  API metadata alone, so these figures are lower bounds.
- Net lines changed is a rough proxy for project size; it includes
  documentation and data files unless filtered out (use the file-level
  output of `gh_commit_lines()` to do so).
- `gh_traffic()` (clones/views) only covers the last 14 days and requires
  push access, so building a long download history requires periodic
  snapshots.
