---
title: "Measuring AI involvement in a repository: the epiworld case"
format:
  html:
    toc: true
    fig-width: 9
    fig-height: 4.5
vignette: >
  %\VignetteIndexEntry{Measuring AI involvement in a repository: the epiworld case}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

This vignette walks through a complete analysis of AI involvement in a real
project: [epiworld](https://github.com/UofUEpiBio/epiworld), a C++
header-only library for agent-based epidemiological simulations developed at
the University of Utah. We will answer three questions:

1. **When** did AI coding agents become involved in the project?
2. **How much** of the commit and coding activity involves AI?
3. **How** did activity change after AI became involved?

```{r}
#| label: setup
library(aitracking)
library(data.table)
```

## Getting the data

Two calls retrieve everything we need. `gh_commits()` downloads the full
commit history, `gh_commit_lines()` adds per-commit line counts (one API call
per commit), and `gh_interactions()` downloads the issue/PR conversation
layer--which survives even when PRs are squash-merged:

```{r}
#| label: fetch
#| eval: false
commits <- gh_commits("UofUEpiBio/epiworld") |>
  gh_commit_lines()

interactions <- gh_interactions("UofUEpiBio/epiworld")
```

Both functions authenticate via `gh_token()` (an explicit token, the
`GITHUB_PAT`/`GITHUB_TOKEN` environment variables, or the git credential
store). So that this vignette builds offline, the package ships a snapshot of
both datasets taken on 2026-07-31, and we use it from here on:

```{r}
#| label: load-data
data(epiworld_commits)
data(epiworld_interactions)

commits      <- epiworld_commits
interactions <- epiworld_interactions

commits[, .(sha = substr(sha, 1, 7), author, date, additions, deletions)] |>
  head() |>
  knitr::kable()
```

## When did AI become involved?

`ai_classify()` tags each commit using a small set of transparent rules: AI
identities among authors/committers (e.g., `copilot[bot]`), and attribution
trailers in the commit message (e.g., `Co-authored-by: Claude`). The default
patterns are in `ai_patterns()`; the generic `bot` pattern also catches
non-AI automation such as Dependabot, so we drop it here to focus on actual
AI coding agents:

```{r}
#| label: classify
ai_only <- ai_patterns()[names(ai_patterns()) != "bot"]

commits <- ai_classify(commits, patterns = ai_only)

# How many commits involved AI, and which agents?
commits[, .N, by = .(ai, ai_agent)] |> knitr::kable()
```

```{r}
#| label: first-ai
first_ai <- commits[ai == TRUE, min(date)]
first_ai
```

So the first AI-involved commit happened on
`r format(first_ai, "%B %d, %Y")`:

```{r}
#| label: first-ai-commit
commits[ai == TRUE][which.min(date), .(
  sha = substr(sha, 1, 7), author, date,
  message = substr(message, 1, 60), ai_agent
)] |> knitr::kable()
```

Note that this commit's *author* is a human. It was flagged because the
message carries an attribution trailer left by a squash-merged pull request:

```{r}
#| label: first-ai-trailer
commits[ai == TRUE][which.min(date), message] |>
  strsplit("\n") |>
  unlist() |>
  grep(pattern = "Co-authored-by", value = TRUE) |>
  unique()
```

This is the main reason `ai_classify()` looks at commit messages and not just
at author identities: when a PR written with an agent is squash-merged, the
resulting commit is attributed to whoever merged it, and the trailer is the
only trace of AI involvement left in the commit history.

## Visualizing the timeline

`plot_timeline()` shows monthly activity. Since the data went through
`ai_classify()`, commit bars are split into human and AI-involved:

```{r}
#| label: timeline-commits
plot_timeline(commits, by = "commits")
```

The same timeline in terms of lines added and deleted needs one adjustment
first. epiworld's recorded git history begins with the import of the
existing codebase, and that import alone lands in a single calendar month:

```{r}
#| label: first-month
first_month       <- format(min(commits$date), "%Y-%m")
first_month_added <- commits[format(date, "%Y-%m") == first_month, sum(additions, na.rm = TRUE)]
rest_added        <- commits[format(date, "%Y-%m") != first_month, sum(additions, na.rm = TRUE)]

commits_since_import <- commits[format(date, "%Y-%m") != first_month]
```

That first month (`r first_month`) alone accounts for
`r format(first_month_added, big.mark = ",")` lines added --
`r round(first_month_added / rest_added, 1)` times as many as every month
since, *combined* (`r format(rest_added, big.mark = ",")`). Leaving it in
would compress the rest of the plot to a flat line near zero, so for this
view only, we drop that first calendar month and start from the second month
of recorded history onward. (Commit *counts* are not skewed the same way --
that first month had 63 commits, well within the normal range -- so the
commit-based timeline above uses the full history unmodified.)

```{r}
#| label: timeline-lines
plot_timeline(commits_since_import, by = "lines")
```

### Zooming in: daily and weekly bins

`interval` also accepts `"week"` and `"day"`. `plot_timeline()` only plots
the range actually present in its input, so to zoom into a shorter window --
say, a specific month or quarter rather than the project's full history --
filter the data first and switch the interval. Here is January through March
2025, the quiet stretch right before AI got involved:

```{r}
#| label: zoom-window
zoom <- commits[date >= as.POSIXct("2025-01-01") & date < as.POSIXct("2025-04-01")]
nrow(zoom)
```

```{r}
#| label: zoom-commits
plot_timeline(zoom, by = "commits", interval = "day")
```

```{r}
#| label: zoom-lines
plot_timeline(zoom, by = "lines", interval = "day")
```

`r nrow(zoom)` commits in three months makes for a sparse plot, but the
binning itself is exactly what you would get on a busier repository or a
wider window -- one bar per calendar day, none of them merged into weeks or
months. The lines view has a spike worth checking rather than taking at face
value:

```{r}
#| label: zoom-spike
zoom[which.max(additions), .(sha = substr(sha, 1, 7), additions, message = substr(message, 1, 60))] |>
  knitr::kable()
```

That commit is adding image assets, not code -- a reminder of the caveat
below that net lines changed includes non-code files. `interval = "week"`
sits between the two, useful when daily bins are too sparse but monthly bins
are too coarse.

## How did activity change since AI got involved?

A simple before/after comparison of monthly activity. The initial-import
month identified above lands entirely in the "before AI" period, and its
million-plus added lines would swamp the comparison, so we reuse
`commits_since_import` here as well:

```{r}
#| label: before-after
commits_since_import[
  , period := fifelse(date < first_ai, "1. Before AI", "2. Since AI")
]

stats <- commits_since_import[, .(
  months        = uniqueN(format(date, "%Y-%m")),
  commits       = .N,
  authors       = uniqueN(author_name),
  lines_added   = sum(additions, na.rm = TRUE),
  lines_deleted = sum(deletions, na.rm = TRUE)
), by = period]

stats[, `:=`(
  commits_month = round(commits / months, 1),
  added_month   = round(lines_added / months, 1),
  deleted_month = round(lines_deleted / months, 1)
)]

stats |> knitr::kable()
```

And within the AI period, how much of the work is AI-involved?

```{r}
#| label: ai-share
share <- commits_since_import[period == "2. Since AI", .(
  commits     = .N,
  lines_added = sum(additions, na.rm = TRUE)
), by = ai]

share |> knitr::kable()
```

```{r}
#| label: share-values
#| include: false
pc_commits <- round(100 * share[ai == TRUE, commits] / sum(share$commits))
pc_lines   <- round(100 * share[ai == TRUE, lines_added] / sum(share$lines_added))
rate_before <- stats[period == "1. Before AI", commits_month]
rate_since  <- stats[period == "2. Since AI", commits_month]
```

Two things stand out. First, AI involvement is substantial once it starts:
`r pc_commits`% of commits and `r pc_lines`% of added lines in the AI period
are AI-involved. Second--and this is why the raw counts deserve care--overall
activity did **not** accelerate: the project went from `r rate_before` to
`r rate_since` commits per month.

That drop is a good reminder that this is observational data, not an
experiment. epiworld is a mature library, and the AI period coincides with a
phase of consolidation rather than initial development; the same comparison
on a young project would likely look very different. What the data supports
is a statement about *composition* (how much of the work AI touches), not
about causation (whether AI made the project faster).

## Human requests to AI: the conversation layer

Commit histories under-count AI involvement: when a PR authored by an agent
is squash-merged, the merge commit may carry a human identity. The
issue/PR conversation retrieved by `gh_interactions()` preserves that
history. `ai_classify()` adds two flags here: `ai` (the comment/issue was
authored by an AI agent) and `ai_mention` (the text addresses or mentions an
agent--a proxy for humans prompting AI):

```{r}
#| label: interactions
interactions <- ai_classify(interactions, patterns = ai_only)

# Overall activity by type
interactions[, .N, by = type] |> knitr::kable()
```

```{r}
#| label: interactions-ai
# Comments/issues authored by AI agents
interactions[ai == TRUE, .N, by = ai_agent] |> knitr::kable()
```

```{r}
#| label: interactions-mentions
# Humans prompting/addressing AI agents
interactions[ai == FALSE & !is.na(ai_mention), .N, by = ai_mention] |>
  knitr::kable()
```

Activity by user (top 10), distinguishing AI agents from humans:

```{r}
#| label: by-user
interactions[, .N, by = .(user, ai)][order(-N)] |>
  head(10) |>
  knitr::kable()
```

## Delegation: what was handed to an agent

Commits and comments tell you where AI *ended up*. Assignments tell you where
it was *asked to go*: assigning an issue to Copilot is how GitHub's coding
agent is given a task. `gh_assignments()` returns one row per issue/assignee
pair and flags the AI ones:

```{r}
#| label: assignments
data(epiworld_assignments)

epiworld_assignments[assigned_ai == TRUE, .N, by = .(type, assignee, assignee_type)] |>
  knitr::kable()
```

```{r}
#| label: assignment-share
epiworld_assignments[, .(
  rows        = .N,
  to_ai       = sum(assigned_ai),
  share_to_ai = round(mean(assigned_ai), 3)
)] |> knitr::kable()
```

Note the `assignee_type`: GitHub reports the coding agent as the login
`Copilot` of type `Bot`, but *without* the usual `[bot]` suffix. Matching on
the suffix alone would miss it entirely -- and matching on `type == "Bot"`
alone would miss agents such as `cursoragent` and `openhands-agent`, which
are plain `User` accounts. This is why detection here is pattern-based.

`plot_timeline()` is not limited to commit tables: pass `date_col` and
`ai_col` to point it at the timestamp and AI-flag columns of a different
table, here `gh_assignments()`'s `created_at` and `assigned_ai`. We drop the
unassigned rows first, since "human vs. AI" is not a meaningful split for an
issue nobody was handed:

```{r}
#| label: assignments-timeline
epiworld_assignments[!is.na(assignee)] |>
  plot_timeline(
    by = "commits", date_col = "created_at", ai_col = "assigned_ai",
    main = "Issues/PRs assigned over time"
  )
```

There is no suspected tier here, and deliberately so: an assignee is a
direct identity, not something inferred from a branch name the way
`ai_suspected` is for commits and pull requests below, so every assignment is
either confirmed-human or confirmed-AI. `plot_timeline()` detects that the
suspected column is absent and drops that band from the legend automatically
-- the same mechanism that would show it if `epiworld_pulls` were plotted
this way instead.

## Confirmed vs. suspected: the branch-prefix signal

Coding agents push to prefixed branches, and that prefix is often the
longest-lived trace of AI involvement: it survives squash-merges and shows up
even where no bot account and no commit trailer do. `gh_pulls()` returns the
head branch, and `ai_classify()` reads it:

```{r}
#| label: pulls
data(epiworld_pulls)

pulls <- ai_classify(epiworld_pulls, patterns = ai_only)

pulls[, .N, by = .(ai, ai_suspected, ai_evidence, ai_agent)] |> knitr::kable()
```

The two tiers are doing real work here. Every `copilot/*` pull request was
opened by the `Copilot` bot account, so it is **confirmed** via `identity`.
The `codex/*` ones look similar but were opened by a human:

```{r}
#| label: suspected
pulls[ai_suspected == TRUE, .(number, user, branch, ai_agent, ai_evidence)] |>
  knitr::kable()
```

That is exactly why a branch prefix is recorded as `ai_suspected` rather than
folded into `ai`. Agents that cannot open pull requests themselves leave this
pattern: a developer runs the agent locally, pushes the branch it produced,
and opens the PR under their own account. The prefix says AI was probably
involved; it does not say who wrote which line, and in principle a human
could name a branch `copilot/...` with no agent involved at all.

Read `ai` as a lower bound on AI involvement and `ai | ai_suspected` as an
upper bound:

```{r}
#| label: bounds
pulls[, .(
  pull_requests = .N,
  confirmed     = sum(ai),
  upper_bound   = sum(ai | ai_suspected)
)] |> knitr::kable()
```

## Evolution of the project's size

`loc_evolution()` accumulates net lines changed (additions minus deletions)
over time--an approximation of the project's size in lines:

```{r}
#| label: loc
lo <- loc_evolution(commits)

plot(
  lo$date, lo$loc, type = "l", lwd = 2, col = "steelblue",
  xlab = "Date", ylab = "Cumulative net lines",
  main = "epiworld: project size over time"
)
abline(v = as.numeric(first_ai), lty = 2, col = "tomato")
text(
  as.numeric(first_ai), max(lo$loc) * 0.1, "first AI commit",
  pos = 4, col = "tomato"
)
```

For a by-language breakdown, retrieve file-level detail with
`gh_commit_lines(commits, files = TRUE)` and pass that to `loc_evolution()`,
or get the current snapshot with:

```{r}
#| label: languages
#| eval: false
gh_languages("UofUEpiBio/epiworld")
```

## Caveats

- The "lines" timeline and the before/after comparison above exclude the
  calendar month in which epiworld's recorded history begins, since it
  contains the initial code import (over a million lines in a single month)
  and would otherwise dominate every other month combined. This exclusion is
  scoped to those two views; the cumulative size plot below is a historical
  record and rightly includes that initial jump.
- `ai_classify()` is deliberately parsimonious: it only catches AI
  involvement that is *visible* (bot identities, attribution trailers,
  mentions). Undisclosed AI assistance--e.g., code written with editor
  completions and committed under a human identity--is not detectable from
  API metadata alone, so these figures are lower bounds.
- Net lines changed is a rough proxy for project size; it includes
  documentation and data files unless filtered out (use the file-level
  output of `gh_commit_lines()` to do so).
- `gh_traffic()` (clones/views) only covers the last 14 days and requires
  push access, so building a long download history requires periodic
  snapshots.
